---
title: "Rethinking Stats"
output: html_document
---
```{r}

# devtools::install_github("stan-dev/cmdstanr")

# cmdstanr::check_cmdstan_toolchain(fix = TRUE)
# cmdstanr::install_cmdstan()

# install.packages(c("coda","mvtnorm","devtools","loo","dagitty"))
# devtools::install_github("rmcelreath/rethinking")
```

##Chapter one

```{r setup, include=FALSE}

grid_size <- 100

source(here::here('R', 'utils.R'))
```

```{r}

calc_posterior( 6 , 9 , grid.size = grid.size )

```

```{r}
prior <- ifelse(p_grid < 0.5, 0, 1)

calc_posterior( 6 , 9 , prior, grid.size=grid.size )

prior <- exp( -5*abs( p_grid - 0.5 ) )

calc_posterior( 6 , 9 , prior, grid.size=grid.size )

calc_posterior(6, 9, prior, grid.size=grid.size)
```

```{r}
#Q 2M1
#(1) W, W, W
calc_posterior( 3 , 3 , prior, grid.size=grid.size )
#(2) W, W, W, L
calc_posterior( 3 , 4 , prior, grid.size=grid.size )
#(3) L, W, W, L, W, W, W
calc_posterior( 5 , 7 , prior, grid.size=grid.size )

```

```{r}
# Q 2M2

prior <- ifelse( p_grid < 0.5 , 0 , 1 )

#(1) W, W, W
calc_posterior( 3 , 3 , prior, grid.size=grid.size )
#(2) W, W, W, L
calc_posterior( 3 , 4 , prior, grid.size=grid.size )
#(3) L, W, W, L, W, W, W
calc_posterior( 5 , 7 , prior, grid.size=grid.size )

# (1) W, W, W
calc_posterior(3, 3, prior, grid.size=grid.size )
# (2) W, W, W, L
calc_posterior(3, 4, prior, grid.size=grid.size )
# (3) L, W, W, L, W, W, W
calc_posterior(5, 7, prior, grid.size=grid.size )
```

```{r}

# define prior
prior <- prior
# compute likelihood at each value in grid
likelihood <- dbinom(1, size = 1, prob = 0.3)
# compute product of likelihood and prior
unstd.posterior <- likelihood * prior
# standardize the posterior, so it sums to 1
posterior <- unstd.posterior / sum(unstd.posterior)
```

##Chapter two

##Chapter three

```{r}
p_grid <- seq(from = 0, to = 1, length.out = 1000)
prior <- rep(1, 1000)
likelihood <- dbinom(6, size = 9, prob = p_grid)
unstd.posterior <- likelihood * prior
posterior <- unstd.posterior / sum(posterior)

sample.size <- 1e4

set.seed(100)
samples <- sample(p_grid, prob = posterior, size = sample.size, replace = TRUE)
```


```{r}
plot(p_grid, posterior)

plot(samples)
```

```{r}
library(rethinking)

dens(samples)
```

## Intervals of defined boundaries

### How much posterior probability lies below some parameter value?

```{r}
# add up posterior probability where p < 0.5
sum(posterior[p_grid < 0.5])

# the same calculation, using samples from the posterior
sum(samples < 0.5) / sample.size
```

### How much posterior probability lies between two parameter values?

```{r}
# how much posterior probability lies between 0.5 and 0.75:

sum(samples > 0.5 & samples < 0.75) / sample.size
```

## Intervals of defined mass

```{r}
# you want to know the boundaries of the lower 80% posterior probability. You know this interval starts at p = 0. To find out where it stops, think of the samples as data and ask where the 80th percentile lies.

quantile(samples, 0.8)

# Similarly, the middle 80% interval lies between the 10th percentile and the 90th percentile. These boundaries are found using the same approach

quantile(samples, c(0.1, 0.9))
```


### 3E1. How much posterior probability lies below p = 0.2?
### 3E2. How much posterior probability lies above p = 0.8?
### 3E3. How much posterior probability lies between p = 0.2 and p = 0.8?

```{r}
`3E1` <- sum(posterior[p_grid < 0.2])
`3E2` <- sum(posterior[p_grid > 0.8])
`3E3` <- sum(posterior[p_grid > 0.2 & p_grid < 0.8])

`3E1` + `3E2` + `3E3`
```

### 3E4. 20% of the posterior probability lies below which value of p?
### 3E5. 20% of the posterior probability lies above which value of p?

```{r}
quantile(samples, 0.2)
quantile(samples, 0.8)
```

### 3E6. Which values of p contain the narrowest interval equal to 66% of the posterior probability?
### 3E7. Which values of p contain 66% of the posterior probability, assuming equal posterior probability both below and above the interval?

```{r}

HPDI(samples, prob = 0.66)

PI(samples, prob = 0.66)
```

### 3M1. Suppose the globe tossing data had turned out to be 8 water in 15 tosses. Construct the posterior distribution, using grid approximation. Use the same flat prior as before.

```{r}
n_water <- 8
n <- 15
grid.size <- 1e4

calc_posterior(n_water, n, rep(1, grid.size), grid.size)
```

### 3M2. Draw 10,000 samples from the grid approximation from above. Then use the samples to calculate the 90% HPDI for p.

```{r}

sample.size <- 1e4

set.seed(100)
samples <- sample(p_grid, prob = posterior, size = sample.size, replace = TRUE)

HPDI(samples, 0.9)
```

### 3M3. Construct a posterior predictive check for this model and data. This means simulate the distribution of samples, averaging over the posterior uncertainty in p. What is the probability of observing 8 water in 15 tosses?

```{r}

w <- rbinom( 1e4 , size=15 , prob=samples )

simplehist( w )

p <- table(w)/1e4

p[9]

```

### 3M4. Using the posterior distribution constructed from the new (8/15) data, now calculate the probability of observing 6 water in 9 tosses.

```{r}

w <- rbinom( 1e4 , size=9 , prob=samples )

simplehist( w )

p <- table(w)/1e4

p[7]

```

### 3M5. Start over at 3M1, but now use a prior that is zero below p = 0.5 and a constant above p = 0.5.This corresponds to prior information that a majority of the Earth’s surface is water. Repeat each problem above and compare the inferences. What difference does the better prior make? If it helps, compare inferences (using both priors) to the true value p = 0.7.

```{r}

n_water <- 8
n <- 15
grid.size <- 1e4

prior <- ifelse( p_grid < 0.5 , 0 , 1 )

calc_posterior( n_water, n, prior, grid.size )

samples <- sample( p_grid, size=grid.size, prob=posterior, replace = T )

HPDI( samples, 0.9 )

# 3M3

w <- rbinom( 1e4 , size=15 , prob=samples )

simplehist( w )

p <- table(w)/1e4

p[9]

# 3M4

w <- rbinom( 1e4 , size=9 , prob=samples )

simplehist( w )

p <- table(w)/1e4

p[7]

```

### 3M6 *

```{r}

quantile( samples , c(0.99, 1) )

PI( samples, prob=0.99 )

```
### 3H1

```{r}

data(homeworkch3) 

n_boys <- 1
n_births <- 1
grid.size <- 1e4

prior <- rep(1,grid.size)

calc_posterior( n_boys, n_births, prior, grid.size )

# the value of one maximises the posterior probability of a birth being a boy

```

### 3H2

```{r}

samples <- sample( p_grid, size=1e4, replace=T, prob=posterior )

HPDI( samples, prob=0.5 )
HPDI( samples, prob=0.89 )
HPDI( samples, prob=0.97 )

```
### 3H3

```{r}

b <- rbinom( 1e4 , size=200 , prob=samples )
simplehist( b )

p <- table(b)/1e4

p

dens(b)

#does the model include 111 boys as a central likely outcome? no, the model is skewed towards 200 boys in 200 births

```



### 3H4

```{r}

sum(birth1)  

b <- rbinom( 1e4 , size=100 , prob=samples )
dens( b )

```

### 3H5 

```{r}



```

